### 2020-10-20 full_version_1013/setup.tex
start creating arxiv version

%We assume that there are multiple datasets that all follow (potentially different) linear models.
%In each dataset, suppose the feature vector is $x = {\Sigma}^{1/2} z$, where $z \in \real^p$ has i.i.d entries with zero mean and unit variance, and the population covariance matrix $\Sigma \in\real^{p \times p}$ is deterministic and positive semidefinite.
Suppose we have $t$ datasets.
For each dataset $i$ from $1$ to $t$, let $n_i$ denote its sample size.
Let $X^{(i)} \in \real^{n_i \times p}$ denote dataset $i$'s feature covariates.
We assume that the label vector $Y^{(i)} \in \real^{n_i}$ for $X^{(i)}$ follows a linear model plus random noise.
We study the standard hard parameter sharing architecture:
a shared feature representation layer $B\in\real^{p\times r}$ for all datasets and a separate output layer $A_i \in \real^r$ for every dataset $i$.
See Figure \ref{fig_intro_arch} for an illustration.
%The width of $B$, denoted by $r$, plays an important role in regularization.
%As observed in Proposition 1 of \citet{WZR20}, if $r \ge t$, there is no regularization effect.
%Hence, we assume that $r < t$ in our study.
%For example, when there are only two tasks, $r = 1$ and $B$ reduces to a vector whereas $W_1, W_2$ become scalars.
We study the following minimization problem:
%	\item Separate each dataset $(X_i, Y_i)$ randomly into a training set $(X_i^{tr}, Y_i^{tr})$ and a validation set $(X_i^{val}, Y_i^{val})$.
%	The size of each set is described below.
%	\item Learn the shared layer $B$: minimize the training loss over $B$ and $W_1, \dots, W_t$, leading to a local minimum of $B$ that depends on $W_1, \dots, W_t$, denoted by $\hat{B} = \hat{B}(W_1, \dots, W_t)$.
%\vspace{-0.1in}
\begin{align}\label{eq_mtl}
			f(A, B) = \sum_{i=1}^t \norm{X^{(i)} B A_i - Y^{(i)}}^2,
\end{align}
%\vspace{-0.05in}
where $A = [A_1, A_2, \dots, A_t] \in \real^{r \times t}$.
Given a solution from minimizing $f(A, B)$, denoted by $(\hat{A}, \hat{B})$  (which we will specify below), let $\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i$ denote the HPS estimator for task $i$.
The critical questions are:
(i) How well does the estimator work? In particular, how does the performance of the estimator scale with sample size?
(ii) For datasets with different sample sizes and covariate shifts, how do they affect the estimator?

\textbf{Main results.}
Our first result (Theorem \ref{thm_many_tasks}) applies to multi-label prediction settings where all datasets have the same features (and sample size), and we want to make several predictions for every input (cf. examples in \citet{hsu2009multi}).
We analyze the global minimizer of $f(A, B)$, and provide a sharp bias-variance decomposition of its (out-of-sample) prediction loss for any task.
This setting is tractable even though in general, $f(A, B)$ is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
%We show that the prediction loss of HPS admits a clean bias-variance decomposition.
Our result implies that when all tasks have the same features but different labels, for any task, HPS helps reduce the task's variance compared to single-task learning but increases bias.

Our second result (Theorem \ref{thm_main_RMT}) applies to two tasks with arbitrarily different sample size ratios and covariate shifts.
While we can no longer characterize $f(A, B)$'s global minimum because of non-convexity, we can still provide a sharp bias-variance tradeoff of any local minimizer's prediction loss for both tasks.
Despite being a simple setting, we observe several non-trivial phenomena by varying sample size ratios and covariate shifts between the two tasks.
See Figure \ref{fig_intro_sample_size_b} for an illustration of the former.
Consequently, using our precise loss estimates, we observe several qualitative properties of HPS for varying dataset properties.
%\begin{enumerate}

\textit{Sample efficiency (Example \ref{ex_same_cov})}:
	One advantage of combining multiple datasets is that the requirement for labeled data reduces compared to single-task learning, a phenomenon that \citet{ZSSGM18} has observed empirically.
	Our results further imply that HPS's sample efficiency depends on model-specific variances across tasks vs. the noise variance and is generally high when the latter is large.

\textit{Sample size ratio (Example \ref{ex_sample_ratio})}: Increasing one task's sample size does not always reduce another task's loss. In a simplified setting, we find that the task loss either decreases first before increasing afterward or decreases monotonically depending on how fast the bias grows. These two trends result from different bias-variance tradeoffs. This result is surprising because previous generalization bounds in multi-task learning typically scale down as all tasks' sample sizes increase, thus do not apply for different sample size ratios.

\textit{Covariate shift (Example \ref{ex_covshift})}: In addition to sample sizes, variance also scales with two datasets' covariate shifts. For a large sample size ratio, HPS's  variance is smallest when there is no covariate shift. Counterintuitively, for a small sample size ratio, having covariate shifts reduces variance through a complementary spectrum. We achieve this result through a novel characterization on the inverse of the sum of two sample covariance matrices with arbitrary covariate shifts. See our discussion of proof techniques below for details.
%\end{enumerate}

@@

% plus an identity matrix,  due to lack of freeness
%\HZ{add why this is challenging}

Finally, we discuss the practical implications of our work.
Our sample size ratio study implies a concrete progressive training procedure that gradually adds more data until performance drops.
For example, in the setting of Figure \ref{fig_intro_sample_size_b}, this procedure will stop right at the minimum of the local basin.
We conduct further studies of this procedure on six text classification datasets and observe that it reduces the computational cost by $65\%$ compared to a standard round-robin training procedure while keeping the average accuracy of all tasks simultaneously.

\textbf{Proof techniques.}
There are two main ideas in our analysis. The proof of our first result uses a geometric intuition that hard parameter sharing finds a ``rank-$r$'' approximation of the datasets.
We carefully keep track of the concentration error between the global minimizer of $f(A, B)$ and its population version (cf. equation \eqref{eq_A_star}).
The proof of our second result is significantly more involved because of different sample sizes and covariate shifts. We show that the inverse of the sum of two sample covariance matrices with arbitrary covariate shifts converges to a deterministic diagonal matrix asymptotically (cf. Theorem \ref{thm_main_RMT}).
We use recently developed techniques from random matrix theory to show a sharp convergence rate.
% to obtain a sharp estimate on the..., which is commonly referred to as the \emph{local law}.
%\HZ{add several sentences on the technical insight}
One limitation of our analysis is that in Example \ref{ex_sample_ratio}, there is an error term that can result in vacuous bounds for very small $n_1$ (cf. equation \eqref{cor_MTL_error}).
We believe our result has provided significant initial insights, and it is an interesting question to tighten our result.

@@

There is a large body of classical and recent works on multi-task learning.
We focus our discussion on theoretical results and refer interested readers to several excellent surveys for general references \cite{PY09,ZY17,V20}.
The early work of \citet{B00,BS03,M06} studied multi-task learning from a theoretical perspective, often using uniform convergence or Rademacher complexity based techniques.
An influential paper by \citet{BBCK10} provides uniform convergence bounds that combine multiple datasets in certain settings.
One limitation of uniform convergence based techniques is that the results often assume that all  tasks have the same sample size, see e.g. \citet{B00,MPR16}.
Moreover, these techniques do not apply to the high-dimensional setting because the results usually require a sample size of at least $p \log p$.

Our proof techniques use the so-called local law of random matrices \cite{erdos2017dynamical}, a recent development in the random matrix theory literature.
In the single-task case, \citet{isotropic} first proved such a local law for sample covariance matrices with isotropic covariance.
\citet{Anisotropic} later extended this result to arbitrary covariance matrices.
%On the other hand, one may derive the asymptotic result in Theorem \ref{thm_main_RMT} with error $\oo(1)$ using the free addition of two independent random matrices in  theory .
These techniques provide almost sharp convergence rates to the asymptotic limit compared to other methods such as free probability \cite{nica2006lectures}.
To the best of our knowledge, we are not aware of any previous results in the multi-task case, even for two tasks (with arbitrary covariate shifts).

The problem we study here is also related to high-dimensional prediction in transfer learning \cite{li2020transfer,bastani2020predicting} and distributed learning \cite{dobriban2018high}.
For example, \citet{li2020transfer} provide minimax-optimal rates to predict a target regression task given multiple sparse regression tasks.
One closely related work is \citet{WZR20}, which studied hard parameter sharing for two linear regression tasks.
However, their results only apply to sample size regimes at least logarithmic factors of dimension.
%\citet{WZR20} (and an earlier work by \citet{KD12}) observed that the shared layer size $r$ in hard parameter sharing plays a critical role of regularization.
%Linear models in multi-task learning have been studied in various settings, including online learning \cite{CCG10,DCSP18}, sparse regression \cite{LPTV09,LPVT11}, and representation learning \cite{BHKL19}.

@@

\textbf{Organizations.}
The rest of this paper is organized as follows.
In Section \ref{sec_same}, we present the bias-variance decomposition for hard parameter sharing.
In Section \ref{sec_diff}, we describe how varying sample sizes and covariate shifts impact hard parameter sharing using random matrix theory.
In Sections \ref{sec_simulation}, we validate our results in simulations.
In Section \ref{sec_conclude}, we summarize our work and discuss future work.
Section \ref{sec_text} describes our study on text classification tasks.
Section \ref{app_proof_error_same_cov}, \ref{appendix RMT}, and \ref{app_iso_cov} present proofs of our results.

\textbf{Notations.}
%Let $\cE \define [\varepsilon_1, \varepsilon_2, \dots, \varepsilon_t] \in \real^{n \times t}$ denote the random noise.
%We can also write $Y = XB^{\star} + \cE$.
%Let $A = [A_1, A_2, \dots, A_t] \in \real^{r\times t}$ be a matrix notation that contains all the output layer parameters.
For an $n\times p$ matrix $X$, let $\lambda_{\min}(X)$ denote its smallest singular value and $\norm{X}$ denote its largest singular value.
Let $\lambda_1(X), \lambda_2(X), \cdots, \lambda_{p\wedge n}(X)$ denote the singular values of $X$ in decreasing order.
Let $X^+$ denote the Moore-Penrose psuedoinverse of $X$.
We refer to random matrices of the form $\frac {X^\top X} n$ as sample covariance matrices.
We say that an event $\Xi$ holds with high probability if the probability that $\Xi$ happens goes to $1$ as $p$ goes to infinity.
%We shall use $\oo(1)$ to mean a small positive quantity that converges to 0 as $p$ goes to infinity.

>>>>

We assume that there are multiple datasets that all follow (potentially different) linear models.
In each dataset, suppose the feature vector is $x = {\Sigma}^{1/2} z$, where $z \in \real^p$ has i.i.d entries with zero mean and unit variance, and the population covariance matrix $\Sigma \in\real^{p \times p}$ is deterministic and positive semidefinite.

We study a hard parameter sharing architecture to learn jointly from multiple datasets.
In this architecture, there is a shared feature representation layer $B\in\real^{p\times r}$ for all tasks and a separate output layer $A_i \in \real^r$ for every task $i$.
Suppose we have $t$ datasets.
For each dataset $i$ from $1$ to $t$, let $n_i$ denote its sample size, $X^{(i)} \in \real^{n_i \times p}$ denote its feature covariates, and $Y^{(i)} \in \real^{n_i}$ denote all the labels.
%The width of $B$, denoted by $r$, plays an important role in regularization.
%As observed in Proposition 1 of \citet{WZR20}, if $r \ge t$, there is no regularization effect.
%Hence, we assume that $r < t$ in our study.
%For example, when there are only two tasks, $r = 1$ and $B$ reduces to a vector whereas $W_1, W_2$ become scalars.
Let $A = [A_1, A_2, \dots, A_t] \in \real^{r \times t}$.
We study the following optimization objective.
%	\item Separate each dataset $(X_i, Y_i)$ randomly into a training set $(X_i^{tr}, Y_i^{tr})$ and a validation set $(X_i^{val}, Y_i^{val})$.
%	The size of each set is described below.
%	\item Learn the shared layer $B$: minimize the training loss over $B$ and $W_1, \dots, W_t$, leading to a local minimum of $B$ that depends on $W_1, \dots, W_t$, denoted by $\hat{B} = \hat{B}(W_1, \dots, W_t)$.
\begin{align}\label{eq_mtl}
			f(A, B) = \sum_{i=1}^t \norm{X^{(i)} B A_i - Y^{(i)}}^2.
\end{align}
Given a solution $(\hat{A}, \hat{B})$ to the above optimization problem, let $\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i$ denote the hard parameter sharing (HPS) estimator for task $i$.
Some critical questions are:
(i) How well does the estimator work? In particular, how does the performance of the estimator scale with sample size?
(ii) For datasets with different sample sizes and covariate shifts, how do they affect the estimator?

\paragraph{Main results.}
Our first result (Theorem \ref{thm_many_tasks}) applies to the multi-label prediction setting where all datasets have the same features, and we want to make multiple predictions (see e.g. \cite{hsu2009multi}).
We analyze the global minimizer of $f(A, B)$, and provide a sharp generalization bound on its (out-of-sample) prediction loss for any task.
This case is tractable even though in general, $f(A, B)$ is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
We show that the prediction loss of hard parameter sharing admits a clean bias-variance decomposition.
Our results imply that hard parameter sharing helps by reducing variance compared to single-task learning, but hurts by increasing bias.
Our second result (Theorem \ref{thm_main_RMT}) applies to two tasks with arbitrarily different sample size ratios and population covariance matrices.
We analyze the local minimizers of $f(A, B)$ and provide a sharp generalization bound on both tasks' prediction loss.
Despite its simplicity, we show several interesting phenomena by varying sample sizes and covariate shifts in this setting.
See Figure \ref{fig_intro_sample_size} for an illustration.

Consequently, using our precise loss estimates, we observe qualitative properties of hard parameter sharing for varying datasets' properties.
\begin{enumerate}
	\item \textit{Sample efficiency (Example \ref{ex_same_cov})}:
	One advantage of combining multiple datasets is that the requirement for labeled data reduces compared to STL, a phenomenon that \citet{ZSSGM18} has observed empirically.
	Our results further imply that HPS's sample efficiency depends on model-specific variance vs. noise variance. It is generally high when the noise variance is large compared to model-specific variance across different tasks.
	\item \textit{Sample size ratio (Example \ref{ex_sample_ratio})}: Increasing one task's sample size does not always help to reduce another task's loss. In a simplified setting, we find that the task loss either decreases first before increasing afterwards, or decreases monotonically depending on how fast the bias increases. These two trends result from different tradeoffs between increasing bias and decreasing variance.
	\item \textit{Covariate shift (Example \ref{ex_covshift})}: In addition to sample sizes, variance also scales with the covariate shift between different datasets. For a large sample size ratio, HPS's  variance is smallest when there is no covariate shift. Counterintuitively, for a small sample size ratio, having covariate shifts reduces variance through a complementary spectrum.
\end{enumerate}

@@

There are two main ideas in our analysis. The proof of our first result uses a geometric intuition that hard parameter sharing finds a ``rank-$r$'' approximation of the datasets.
We carefully keep track of the concentration error between the global minimum of $f(A, B)$ and its population version.
The proof of our second result is significantly more involved because of different sample sizes and covariate shifts. Using recently developed techniques from random matrix theory \cite{Anisotropic}, we show that the inverse of the sum of two sample covariance matrices with arbitrary covariate shifts converges to a deterministic diagonal matrix asymptotically (cf. Theorem \ref{thm_main_RMT}(i)). % Moreover, we can obtain a sharp bound on the concentration error.
% to obtain a sharp estimate on the..., which is commonly referred to as the \emph{local law}. 
%\HZ{add several sentences on the technical insight} 
One limitation of our analysis is that for two tasks with different sample sizes, there is an extra error term in the prediction loss of hard parameter sharing (cf. equation \eqref{cor_MTL_error}), which can be large for very small $n_1$. This requires studying the spectrum of non-symmetric random matrices, which is an intriguing open question (see Section \ref{sec_conclude} for more discussion).
% plus an identity matrix,  due to lack of freeness
%\HZ{add why this is challenging}

Finally, we discuss the practical implications of work.
Our sample size ratio study implies a concrete progressive training procedure that gradually adds more data until performance drops.
For example, in the setting of Figure \ref{fig_intro_sample_size_b}, this procedure will stop right at the minimum of the local basin.
We conduct further studies of this procedure on six text classification datasets and observe that it reduces the computational cost by $65\%$ compared to a standard round-robin training procedure while keeping the average accuracy of all tasks simultaneously.

@@

There is a large body of both classical and recent works on multi-task learning.
We focus our discussion on theoretical works, and refer interested readers to several excellent surveys for general references \cite{PY09,R17,ZY17,V20}.
The early work of \citet{B00,BS03,M06} have sought for a study of multi-task learning from a theoretical perspective, often using uniform convergence or Rademacher complexity based techniques.
An influential paper by \citet{BBCK10} provides uniform convergence bound that combines multiple datasets in certain settings.
One limitation of uniform convergence based techniques is that the results often assume that all  tasks have the same sample size, see e.g. \citet{B00,MPR16}.
Moreover, these techniques do not apply to the high-dimensional setting, because the results usually require a sample size at least $p \log p$.

Our proof techniques use the so-called local law of random matrices \cite{erdos2017dynamical}, which is a recent development in the random matrix theory literature.
\citet{isotropic} first proved such a local law for sample covariance matrices with isotropic covariance.
\citet{Anisotropic} later extended this result to arbitrary covariance setting.
%On the other hand, one may derive the asymptotic result in Theorem \ref{thm_main_RMT} with error $\oo(1)$ using the free addition of two independent random matrices in  theory .
These techniques provide almost sharp convergence rates to the asymptotic limit compared to other techniques such as free probability \cite{nica2006lectures}.
To the best of our knowledge, we are not aware of any previous results about the inverse of the sum of two sample covariance matrices with arbitrary covariate shifts.

The problem we study here is also related to high-dimensional prediction in transfer learning \cite{li2020transfer,bastani2020predicting} and distributed learning \cite{dobriban2018high}.
For example, \citet{li2020transfer} provides minimax optimal rates for predicting a target regression task given multiple sparse regression tasks.
One closely related work is \citet{WZR20}, which studied hard parameter sharing for two linear regression tasks.
\citet{WZR20} (and an earlier work by \citet{KD12}) observed that the shared layer size $r$ in hard parameter sharing plays a critical role of regularization.
%Linear models in multi-task learning have been studied in various settings, including online learning \cite{CCG10,DCSP18}, sparse regression \cite{LPTV09,LPVT11}, and representation learning \cite{BHKL19}.

@@

\smallskip
\noindent\textbf{Organizations.}
The rest of this paper is organized as follows.
In Section \ref{sec_same}, we present the bias-variance decomposition for hard parameter sharing.
In Section \ref{sec_diff}, we present our technical results that describe how varying sample sizes and covariate shifts impact hard parameter sharing using random matrix theory.
In Sections \ref{sec_simulation} and \ref{sec_text}, we validate our theory in both simulations and a real world classification task.
In Section \ref{sec_conclude}, we conclude the paper and describe several open questions.

\paragraph{Notations.}
%Let $\cE \define [\varepsilon_1, \varepsilon_2, \dots, \varepsilon_t] \in \real^{n \times t}$ denote the random noise.
%We can also write $Y = XB^{\star} + \cE$.
%Let $A = [A_1, A_2, \dots, A_t] \in \real^{r\times t}$ be a matrix notation that contains all the output layer parameters.
For an $n\times p$ matrix $X$, let $\lambda_{\min}(X)$ denote its smallest singular value and $\norm{X}$ denote its spectral norm.
Let $\lambda_1(X) \ge \lambda_2(X) \ge \cdots \ge \lambda_{p\wedge n}(X)$ denote the singular values of $X$.
Let $X^+$ denote the Moore-Penrose psuedoinverse.
We refer to random matrices of the form $\frac {X^\top X} n$ as sample covariance matrices.
We say that an event $\Xi$ holds with high probability if the probability that $\Xi$ happens goes to $1$ as $p$ goes to infinity.
%We shall use $\oo(1)$ to mean a small positive quantity that converges to 0 as $p$ goes to infinity.
